### 2021-07-03 docs/manuscript/manuscript.tex
Change order of supplementary figures

The phenotype of a GRN was defined as the expression pattern of each gene along
the morphogenetic field after 500 time steps of integration of the dynamical
system. We chose that number of time steps because numerical experiments showed
that GRNs reached the steady state in approximately 300 steps (\nameref{S2_Fig}).

@@

In order to classify each topology by its subgraph composition, we first created
a matrix  $T = (t_{ij})$ with each row corresponding to a different topology
and 17 columns corresponding to subgraphs shown in \nameref{S4_Fig}.
Each element $t_{ij}$ of
the matrix was 1 if the \emph{j}-th subgraph was present in the \emph{i}-th
topology or 0 if the subgraph was not present.\\

@@

\subsection*{Change in morphogenetic field size}
We wanted to evaluate if the GRNs evolved in our model with a morphogenetic
field size of 30 cells could produce the desired pattern in morphogenetic fields
with a different number of cells. For this reason, we calculated the fitness of
each GRN in morphogenetic fields with 10, 20, 40, and 50 cells. In these
simulations we redefined the morphogen gradient in such a way that the initial
and final concentration were constant in all the fields (\nameref{S5_Fig}A).
In order to know if topologies with the same topology had a similar fitness
in the different fields, we performed a Principal Component Analysis in
Scikit-learn~0.24.2~\cite{sklearn} using the fitness values of each GRN in
the four different fields.

@@

Besides topological robustness, we wanted to know if network topologies were
robust to perturbations in the values of the parameters. After performing the
perturbations to a GRN from each topology, we plotted the abundance of the
topology vs. its robustness. We did not find a significant correlation between
these two variables ($r_S = −0.0389724$, $\text{p-value} = 0.298475$)
(Fig~\ref{fig:ab-rob}A). However, when we grouped those topologies containing
the same subgraph and calculated their mean robustness, we found that between
the most robust subgraphs is the ``Bistable'' motif reported by Cotterell \&
Sharpe~\cite{Cotterell2010}, the I3-FFL and the ``Overlapping domains'' motif
(which is a variation of the I3-FFL) also reported by Cotterell \& Sharpe
(see \nameref{S4_Fig} and \nameref{S1_Table}).

@@

The average vertex degree indicates that in general, all the network topologies
display high robustness and that they could resist an average of approximately 4
topological mutations without losing their fitness. Additionally, the average
path length indicates that the evolutionary landscape is easily accessible for
most network topologies (Table~\ref{table1}). Although the GRNs in this modelling
framework are robust to changes in its topology, they are sensitive to changes in
the morphogenetic field size, especially when this size is reduced, as only
4.66, 56.82, 56.14 and 41.05 percent of the GRNs had a fitness greater than 0.9
in fields with 10, 20, 40 and 50 cells, respectively (\nameref{S5_Fig}B).

@@

Hierarchical clustering of spatiotemporal expression profiles showed that GRNs
with topologies that seemed to be very different have indeed very similar
expression dynamics. We expected to find GRNs with the same topology to group
together but instead we found GRNs with different topologies grouped as close
neighbors (Fig~\ref{fig:clustering}A). More interestingly, some GRNs displaying
very abundant network topologies were grouped together with GRNs displaying
very infrequent network topologies, indicating that additional interactions
in the networks have often the function of fine tunning gene expression, instead
of drastically changing the expression mechanism. This can be seen in the
expression dynamics of networks with different topologies (\nameref{S7_Fig}).

@@

These two events conduce to the repression of the gene C by the gene A in the
rightmost part of the morphogenetic field and by the morphogen in the leftmost
part of the morphogenetic field, thus generating a slightly higher level of
expression of gene C in the center of the field. Because gene C activates its
own expression, a band of expression begins to form. This band of expression
then begins to activate the gene A also in the center of the field
(\nameref{S1_Fig}).
% (Fig~\ref{fig:topol9-dyn}).

@@

Robustness is an important feature of patterning mediated by morphogens, and
the GRNs involved can be selected in an evolutionary process not only by their final
steady state but also by their ability to buffer noise~\cite{lo_robust_2015, exelby_2021}.
In our study, the most abundant topologies were
found to be part of highly connected clusters in the neutral network, and this fact
indicates that these topologies are both robust to quantitative changes in the
strength of interactions and diffusion and degradation parameters as well as to
qualitative changes in the topology itself. This, in principle, means that many
mechanisms for producing striped patterns are available and accessible to each
other, i.e., an organism can change between different GRNs (for example between
a Gierer-Meinhardt-like network to an Incoherent Feed-Forward Loop) without losing its
fitness, which allows to fine-tune the regulatory system according to its specific
needs. As robustness and evolvability are
not mutually exclusive properties~\cite{wagner_robustness_2008}, it is likely
that these topologies are also evolvable in the sense that they can produce
heritable variation of the expression pattern.\\

In contrast to this robustness to
qualitative and quantitative changes, we found that many GRNs were susceptible
to changes in morphogenetic field size, but interestingly, GRNs with the same
topology behave in different ways for each field size (\nameref{S5_Fig}C),
indicating that the specific values of network parameters are more important
than the network topology when it comes to producing the same pattern at
different scales.

@@

However, as our model was less stringent and allowed the morphogen input to
interact with any gene, we found a larger number of possible topologies with the
same striped phenotype and a diverse combination of diffusion and degradation
parameters (\nameref{S6_Fig}).
For example, in a single input model the I4-FFL could
not produce a stripe of gene expression~\cite{munteanu_2014}, whereas in our
experimental setup 20.17\% of the topologies presented this motif as a
subgraph~(\nameref{S1_Table}).

@@

\paragraph*{S1 Fig.}
\label{S1_Fig}
{\bf Dynamics of expression of topology number 9.}
The dynamics of expression is shown from $t = 5$ to $t = 150$. The red line
represents the expression level of the gene C along the morphogenetic field,
whereas the dotted line represents the expression level of gene A. The
striped pattern of gene expression can be seen for both genes since $t = 80$.

\paragraph*{S2 Fig.}
\label{S2_Fig}
{\bf Most GRNs reach the steady state before 250 time steps.}
(A) Gene regulatory network displaying the topology number 33. (B) Spatiotemporal
expression profile of the gene regulatory network shown in (A).

\paragraph*{S3 Fig.}
\label{S3_Fig}
{\bf Bridge topologies in the neutral network.}
``Bridge topologies'' are those topologies that connect two clusters in the neutral
network (Fig~\ref{fig:neutral-network}), and as such can be interpreted as
intermediary steps in the evolution from a cluster of related topologies into
another cluster. For example the topology 441 could be an initial step to reach the
topology 9, as in this topology nodes B and C are not connected.

\paragraph*{S4 Fig.}
\label{S4_Fig}
{\bf Relevant subgraphs present in the set of GRNs.}
These are subgraphs that have been reported in previous studies as networks
involved in morphogenesis and development~\cite{Cotterell2010, Schaerli2014,
Schaerli2018}. These were used to calculate the subgraph profile and the results
reported in \nameref{S1_Table}.

\paragraph*{S5 Fig.}
\label{S5_Fig}
{\bf Response of GRNs to changes in morphogenetic field size.}
(A) Morphogen gradients of fields with 10, 20, 40 and 50 cells. (B) Fitness of the
GRNs in each one of the morphogenetic fields. (C) Principal Component Analysis
of fitness by morphogenetic field size. Although principal components separate
GRNs in clusters, not all the GRNs with the same topology are located in the
same group. (D) Average fitness of GRNs evolved in different morphogenetic
field sizes.

\paragraph*{S6 Fig.}
\label{S6_Fig}
{\bf Diffusion and degradation parameters.}
(A) Ternary plot showing the combination of diffusion parameters for genes A,
B and C. (B) Ternary plot showing the combination of degradation parameters
for genes A, B and C.

\paragraph*{S7 Fig.}
\label{S7_Fig}
{\bf Expression dynamics of main networks.}

\paragraph*{S1 Table.}
\label{S1_Table}
{\bf Proportion of relevant subgraphs in the set of GRNs.}
The 19 subgraphs presented in this table are those presented in \nameref{S4_Fig}.
Most of the topologies presented at least one negative feedback loop (70.73\%),
and 62.46\% of them presented the I3-FFL network motif. The robustness of each
subgraph was calculated as the mean robustness of the topologies presenting that
subgraph.

>>>>

The phenotype of a GRN was defined as the expression pattern of each gene along
the morphogenetic field after 500 time steps of integration of the dynamical
system. We chose that number of time steps because numerical experiments showed
that GRNs reached the steady state in approximately 300 steps (\nameref{S1_Fig}).

@@

In order to classify each topology by its subgraph composition, we first created
a matrix  $T = (t_{ij})$ with each row corresponding to a different topology
and 17 columns corresponding to subgraphs shown in \nameref{S2_Fig}.
Each element $t_{ij}$ of
the matrix was 1 if the \emph{j}-th subgraph was present in the \emph{i}-th
topology or 0 if the subgraph was not present.\\

@@

\subsection*{Change in morphogenetic field size}
We wanted to evaluate if the GRNs evolved in our model with a morphogenetic
field size of 30 cells could produce the desired pattern in morphogenetic fields
with a different number of cells. For this reason, we calculated the fitness of
each GRN in morphogenetic fields with 10, 20, 40, and 50 cells. In these
simulations we redefined the morphogen gradient in such a way that the initial
and final concentration were constant in all the fields (\nameref{S3_Fig}A).
In order to know if topologies with the same topology had a similar fitness
in the different fields, we performed a Principal Component Analysis in
Scikit-learn~0.24.2~\cite{sklearn} using the fitness values of each GRN in
the four different fields.

@@

Besides topological robustness, we wanted to know if network topologies were
robust to perturbations in the values of the parameters. After performing the
perturbations to a GRN from each topology, we plotted the abundance of the
topology vs. its robustness. We did not find a significant correlation between
these two variables ($r_S = −0.0389724$, $\text{p-value} = 0.298475$)
(Fig~\ref{fig:ab-rob}A). However, when we grouped those topologies containing
the same subgraph and calculated their mean robustness, we found that between
the most robust subgraphs is the ``Bistable'' motif reported by Cotterell \&
Sharpe~\cite{Cotterell2010}, the I3-FFL and the ``Overlapping domains'' motif
(which is a variation of the I3-FFL) also reported by Cotterell \& Sharpe
(see \nameref{S2_Fig} and \nameref{S1_Table}).

@@

The average vertex degree indicates that in general, all the network topologies
display high robustness and that they could resist an average of approximately 4
topological mutations without losing their fitness. Additionally, the average
path length indicates that the evolutionary landscape is easily accessible for
most network topologies (Table~\ref{table1}). Although the GRNs in this modelling
framework are robust to changes in its topology, they are sensitive to changes in
the morphogenetic field size, especially when this size is reduced, as only
4.66, 56.82, 56.14 and 41.05 percent of the GRNs had a fitness greater than 0.9
in fields with 10, 20, 40 and 50 cells, respectively (\nameref{S3_Fig}B).

@@

Hierarchical clustering of spatiotemporal expression profiles showed that GRNs
with topologies that seemed to be very different have indeed very similar
expression dynamics. We expected to find GRNs with the same topology to group
together but instead we found GRNs with different topologies grouped as close
neighbors (Fig~\ref{fig:clustering}A). More interestingly, some GRNs displaying
very abundant network topologies were grouped together with GRNs displaying
very infrequent network topologies, indicating that additional interactions
in the networks have often the function of fine tunning gene expression, instead
of drastically changing the expression mechanism. This can be seen in the
expression dynamics of networks with different topologies (\nameref{S4_Fig}).

@@

These two events conduce to the repression of the gene C by the gene A in the
rightmost part of the morphogenetic field and by the morphogen in the leftmost
part of the morphogenetic field, thus generating a slightly higher level of
expression of gene C in the center of the field. Because gene C activates its
own expression, a band of expression begins to form. This band of expression
then begins to activate the gene A also in the center of the field
(\nameref{S5_Fig}).
% (Fig~\ref{fig:topol9-dyn}).

@@

Robustness is an important feature of patterning mediated by morphogens, and
the GRNs involved can be selected in an evolutionary process not only by their final
steady state but also by their ability to buffer noise~\cite{lo_robust_2015, exelby_2021}.
In our study, the most abundant topologies were
found to be part of highly connected clusters in the neutral network, and this fact
indicates that these topologies are both robust to quantitative changes in the
strength of interactions and diffusion and degradation parameters as well as to
qualitative changes in the topology itself. This, in principle, means that many
mechanisms for producing striped patterns are available and accessible to each
other through bridge topologies (\nameref{S6_Fig}), i.e.,
an organism can change between different GRNs (for example between
a Gierer-Meinhardt-like network to an Incoherent Feed-Forward Loop) without losing its
fitness, which allows to fine-tune the regulatory system according to its specific
needs. As robustness and evolvability are
not mutually exclusive properties~\cite{wagner_robustness_2008}, it is likely
that these topologies are also evolvable in the sense that they can produce
heritable variation of the expression pattern.\\

In contrast to this robustness to
qualitative and quantitative changes, we found that many GRNs were susceptible
to changes in morphogenetic field size, but interestingly, GRNs with the same
topology behave in different ways for each field size (\nameref{S3_Fig}C),
indicating that the specific values of network parameters are more important
than the network topology when it comes to producing the same pattern at
different scales.

@@

However, as our model was less stringent and allowed the morphogen input to
interact with any gene, we found a larger number of possible topologies with the
same striped phenotype and a diverse combination of diffusion and degradation
parameters (\nameref{S7_Fig}).
For example, in a single input model the I4-FFL could
not produce a stripe of gene expression~\cite{munteanu_2014}, whereas in our
experimental setup 20.17\% of the topologies presented this motif as a
subgraph~(\nameref{S1_Table}).

@@

\paragraph*{S1 Fig.}
\label{S1_Fig}
{\bf Most GRNs reach the steady state before 250 time steps.}
(A) Gene regulatory network displaying the topology number 33. (B) Spatiotemporal
expression profile of the gene regulatory network shown in (A).

\paragraph*{S2 Fig.}
\label{S2_Fig}
{\bf Relevant subgraphs present in the set of GRNs.}
These are subgraphs that have been reported in previous studies as networks
involved in morphogenesis and development~\cite{Cotterell2010, Schaerli2014,
Schaerli2018}. These were used to calculate the subgraph profile and the results
reported in \nameref{S1_Table}.

\paragraph*{S3 Fig.}
\label{S3_Fig}
{\bf Response of GRNs to changes in morphogenetic field size.}
(A) Morphogen gradients of fields with 10, 20, 40 and 50 cells. (B) Fitness of the
GRNs in each one of the morphogenetic fields. (C) Principal Component Analysis
of fitness by morphogenetic field size. Although principal components separate
GRNs in clusters, not all the GRNs with the same topology are located in the
same group. (D) Average fitness of GRNs evolved in different morphogenetic
field sizes.

\paragraph*{S4 Fig.}
\label{S4_Fig}
{\bf Expression dynamics of main networks.}

\paragraph*{S5 Fig.}
\label{S5_Fig}
{\bf Dynamics of expression of topology number 9.}
The dynamics of expression is shown from $t = 5$ to $t = 150$. The red line
represents the expression level of the gene C along the morphogenetic field,
whereas the dotted line represents the expression level of gene A. The
striped pattern of gene expression can be seen for both genes since $t = 80$.

\paragraph*{S6 Fig.}
\label{S6_Fig}
{\bf Bridge topologies in the neutral network.}
``Bridge topologies'' are those topologies that connect two clusters in the neutral
network (Fig~\ref{fig:neutral-network}), and as such can be interpreted as
intermediary steps in the evolution from a cluster of related topologies into
another cluster. For example the topology 441 could be an initial step to reach the
topology 9, as in this topology nodes B and C are not connected.

\paragraph*{S7 Fig.}
\label{S7_Fig}
{\bf Diffusion and degradation parameters.}
(A) Ternary plot showing the combination of diffusion parameters for genes A,
B and C. (B) Ternary plot showing the combination of degradation parameters
for genes A, B and C.

\paragraph*{S1 Table.}
\label{S1_Table}
{\bf Proportion of relevant subgraphs in the set of GRNs.}
The 19 subgraphs presented in this table are those presented in \nameref{S2_Fig}.
Most of the topologies presented at least one negative feedback loop (70.73\%),
and 62.46\% of them presented the I3-FFL network motif. The robustness of each
subgraph was calculated as the mean robustness of the topologies presenting that
subgraph.
